Covariate analysis of ROSMAP reprocessed counts (with CQN normalisation and covariates adjusted)

### 2017-03-15 code/Rmd/ROSMAP_REPROCESSED.Rmd
Minor bug fix in ROSMAP reprocessing code

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering, cache=TRUE}
#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv", "PF_INDEL_RATE", "PF_MISMATCH_RATE"
primaryVariable <- 'cogdx'
FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
ContCovariates <- c("RINcontinuous","age_first_ad_dx","age_death","age_at_visit_max","pmi","educ", 
                    "PCT_PF_READS", "PCT_PF_READS_ALIGNED", "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", 
                    "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")

@@

# Add in RIN^2 values
COVARIATES$RINcontinuous2 = COVARIATES$RINcontinuous^2
ContCovariates = c(ContCovariates, 'RINcontinuous2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=10, fig.height=10}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
```

```{r dataExplore, fig.width = 10, fig.height = 14}
# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = cogdx, y = RINcontinuous)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')

@@

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
p <- p + theme_bw() + theme(legend.position="right")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', primaryVariable)])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preAdjusted.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative design)
Since many covariates are correlated, re-normalising and re-adjusting COUNTS with an iterative design matrix. Adding Batch and msex a priori to variable selection
```{r customNorm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Batch', 'msex');

# Assign residual covariates
residualCovars = setdiff(preAdjustedSigCovars$significantCovars, c(postAdjustCovars, primaryVariable))
covariatesEffects = preAdjustedSigCovars$Effects.significantCovars[residualCovars]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

@@

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

@@

### Clustering residual data
```{r decompse.normalise.data2, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

@@

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', primaryVariable)])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=6, fig.width=18}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable, postAdjustCovars), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

@@

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de}
pl = list()
pl[[1]] = ggplot(DE, aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point()
pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE, aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

@@

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('ROSMAP_DLPFC_offset.tsv', name = 'Gene length and GC content offset)', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

@@

# Store surrogate variables
SURR.VAR %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'ROSMAP_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
SURR_OBJ = File('ROSMAP_DLPFC_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
annotations(SURR_OBJ) = annotations(CODE)
annotations(SURR_OBJ)$dataSubType = 'surrogateVariables'
SURR_OBJ = synStore(SURR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for eQTL analysis
RESIDUAL.SVA.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
eEXP_OBJ = File('ROSMAP_DLPFC_eQTLResidualExpression.tsv', 
                name = 'Normalised, Covariates and SVA Adjusted Residual Expression (for eQTL)', 
               parentId = CODE$properties$id)
annotations(eEXP_OBJ) = annotations(CODE)
annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTL'
eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for eQTL analysis
RESIDUAL.NET.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('ROSMAP_DLPFC_netResidualExpression.tsv', 
                name = 'Normalised, Covariates Removed Residual Expression (for network analysis)', 
               parentId = CODE$properties$id)
annotations(nEXP_OBJ) = annotations(CODE)
annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store differential expression results
write.table(DE, file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('ROSMAP_DLPFC_DiffExpression.tsv', name = 'Differential Expression Results (cogdx)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene lengeth and gc content)   |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |

>>>>

### Covariate clustering
Determine relationship between covariates
```{r covariates.clustering, cache=TRUE}
#"braaksc","ceradsc","cts_mmse30_first_ad_dx","cts_mmse30_lv", "PF_INDEL_RATE", "PF_MISMATCH_RATE", "age_first_ad_dx","age_at_visit_max",
primaryVariable <- "cogdx"
FactorCovariates <- c("Batch", "msex", "race", "spanish", "cogdx")
ContCovariates <- c("RINcontinuous", "age_death", "pmi", "educ", "PCT_PF_READS", "PCT_PF_READS_ALIGNED",
                    "PCT_INTERGENIC_BASES", "PCT_INTRONIC_BASES", "PCT_RIBOSOMAL_BASES", "PCT_USABLE_BASES")

@@

# Add in RIN^2 values
COVARIATES$RINcontinuous2 = COVARIATES$RINcontinuous^2
ContCovariates = c(ContCovariates, 'RINcontinuous2')
```
Correlation/association between covariates at an FDR <= 0.1
```{r covariates.correlation, fig.width=6, fig.height=6}
COVARIATES.CORRELATION = getAssociationStatistics(COVARIATES, PVAL = 0.05)
draw(COVARIATES.CORRELATION$plot, heatmap_legend_side = 'left')
```

```{r data.explore, fig.width = 10, fig.height = 14}
# RIN
p = list()
p[[1]] = ggplot(COVARIATES, aes(x = cogdx, y = RINcontinuous)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')

@@

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=6, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Batch, shape=cogdx, size=RINcontinuous))
p <- p + theme_bw() + theme(legend.position="right")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based classification of samples
```{r decompse.normalise.data1.2, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', primaryVariable)])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates is used to find significant covariates
```{r preadj.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preadj.covariates.plot, fig.width=12, fig.height=8}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative design)
Since many covariates are correlated, re-normalising and re-adjusting COUNTS with an iterative design matrix. Adding Batch and msex a priori to variable selection
```{r iterative.norm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Batch', 'msex');

# Assign residual covariates
residualCovars = setdiff(preAdjustedSigCovars$significantCovars, c(postAdjustCovars, primaryVariable))
residualSigCovars = preAdjustedSigCovars
covariatesEffects = preAdjustedSigCovars$Effects.significantCovars[residualCovars]
postAdjustCovars = c(postAdjustCovars, names(which.max(covariatesEffects))) %>% unique()

@@

### Sanity check
```{r residual.adj, fig.width=12, fig.height=8}
# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)

@@

### Clustering residual data
```{r decompse.normalise.data2.1, fig.height=6, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T, center = T)

@@

```{r decompse.normalise.data2.2, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Batch', 'msex', primaryVariable)])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

Genes that are differentially expressed at an FDR <= 0.05 are
```{r diff.exp, fig.height=6, fig.width=18}
# Get design matrix
DESIGN = getDesignMatrix(COVARIATES[, c(primaryVariable, postAdjustCovars), drop = F], Intercept = F)
DESIGN$design = DESIGN$design[,linColumnFinder(DESIGN$design)$indepCols]

@@

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height=10, fig.width=15}
pl = list()
pl[[1]] = ggplot(DE, aes(x = log10(gene.length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(gene.length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE, aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE, aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(Comparison~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

@@

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('ROSMAP_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

@@

# Store surrogate variables
SURR.VAR %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'ROSMAP_DLPFC_SurVar.tsv', sep = '\t', row.names=F, quote=F)
SURR_OBJ = File('ROSMAP_DLPFC_SurVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
annotations(SURR_OBJ) = annotations(CODE)
annotations(SURR_OBJ)$dataSubType = 'surrogateVariables'
SURR_OBJ = synStore(SURR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for eQTL analysis
RESIDUAL.SVA.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_eQTLResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
eEXP_OBJ = File('ROSMAP_DLPFC_eQTLResidualExpression.tsv', 
                name = 'Normalised, covariates and surrogate variable adjusted residual expression (for eQTL)',
                parentId = CODE$properties$id)
annotations(eEXP_OBJ) = annotations(CODE)
annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForeQTL'
eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for eQTL analysis
RESIDUAL.NET.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('ROSMAP_DLPFC_netResidualExpression.tsv', 
                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
annotations(nEXP_OBJ) = annotations(CODE)
annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store differential expression results
write.table(DE, file = 'ROSMAP_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('ROSMAP_DLPFC_DiffExpression.tsv', name = 'Differential Expression Results (cogdx)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)   |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR_OBJ$properties$id`      |
|  Residual Expression (for eQTL)             |  `r eEXP_OBJ$properties$id`      |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
